--- 
title: "Introduction and Selection Effects" 
author: "Stijn Masschelein" 
date: "`r Sys.Date()`" 
output: rmarkdown::html_vignette 
vignette: > 
  %\VignetteIndexEntry{My Use Case} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8} 
---

## Selection Pressure and Complementarity

Under ideal circumstance, where the complement drives most of the performance
($\beta_1 = \beta_2 = 0, sd = 0.5$), the matching approach is superior whenever
selection pressure is at non-neglible levels. The matching method is able to
detect the complementarity when using a significance level of 0.05. The actual
false detection rate is given by the simulations with $\beta_{12} = 0$ while the
power is given by the simulations with $\beta_{12} = 1$. The traditional
interaction approach quickly loses power to detect the complementarity effect 
when the survival pressure increases. The augmented approach dominates the
traditional method over all rate of survival levels. When diminishing returns 
are an acceptable assumption, the power and error rate control of the intraction
approach can be improved by adding the quadratic terms $x_1^2$ and $x_2^2$.

```{r cache=TRUE, fig.width = 7, fig.align = 'center'}
devtools::load_all(".")
require(dplyr)
require(ggplot2)
require(ggthemes)
rates <- list(0.95, 0.75, 0.5, 0.25, 0.1, 0.05)
bs <- list(c(0, 0, 0, 0), c(0, 0, 0, 1))
nobs <- 100
nsim <- 500
sim1 <- run_sim(nsim = nsim, obs = nobs, rate = rates, b = bs) %>% tbl_df()

sim1_summ <- group_by(sim1, method, b, survival_rate) %>% 
  summarise(detect = mean(pvalue < .05 & coefficient > 0))

sim1_plot <- ggplot(sim1_summ, aes(y = detect, 
                                   x = as.factor(survival_rate), 
                                   colour = b)) + 
  geom_point() +
  geom_hline(yintercept = .05, colour = "grey") +
  geom_hline(yintercept = .025, colour = "grey") +
  facet_wrap( ~ method) + 
  theme_tufte()

print(sim1_plot)  
```

When increasing the noise factor, both methods remain to have a reasonable error
control but power drops dramatically. The matching approach remains to have 
reasonable power as long as the survival pressure is sufficiently high. The
augmented interaction appproach looses most of its shine compared to the
traditional interaction approach when noise increases.

```{r cache=TRUE, fig.width = 7, fig.align = 'center'}
rates <- list(0.75, 0.5, 0.25, 0.1)
sds <- list(0.5, 1, 2, 4)
bs <- list(c(0, 0, 0, 0), c(0, 0, 0, 1))
nobs <- 100
nsim <- 500
sim2 <- run_sim(nsim = nsim, obs = nobs, rate = rates, b = bs, sd = sds) %>% tbl_df()

sim2_summ <- group_by(sim2, method, b, sd, survival_rate) %>% 
  summarise(detect = mean(pvalue < .05 & coefficient > 0))

sim2_plot <- ggplot(sim2_summ, aes(y = detect, 
                                   x = as.factor(survival_rate), 
                                   colour = log(sd), shape = b)) + 
  geom_point(alpha = .5) +
  geom_hline(yintercept = .05, colour = "grey") +
  geom_hline(yintercept = .025, colour = "grey") +
  facet_wrap( ~ method) + 
  theme_tufte()
```

To get an idea about the reason for these differences, I plot the estimated 
coefficient and the $R^2$ for each test. The spread of the coefficients in the
first figure is less pronouned for the matching period than for the interaction 
method. The coefficients are attenuated for both methods depending on the 
survival rate and nois. The combined attenuation is stronger for the interaction
method than for the matching period.

The $R^2$ is not strongly affected by the survival rate and noise except in the 
ideal situation where noise is low and the selection pressure is low. The $R^2$ 
is generally higher for the matching approach. One possible explanation is that
the effect of the noise is on performance, $y$, which is not included in the
matching model.

```{r fig.align='center', fig.width=7}
sim2_plot_coef <- ggplot(filter(sim2, b == "0, 0, 0, 1", 
                                method != "interaction_traditional"), 
                         aes(y = coefficient, x = as.factor(survival_rate))) +
  geom_boxplot() +
  facet_grid(~ method + sd) +
  theme_tufte()

sim2_plot_r <- ggplot(filter(sim2, b == "0, 0, 0, 1", 
                                method != "interaction_traditional"), 
                         aes(y = r2, x = as.factor(survival_rate))) +
  geom_boxplot() +
  facet_grid(~ method + sd) +
  theme_tufte()

plot(sim2_plot_coef)
plot(sim2_plot_r)
```


## 





